---
title: "Namibian hake model update, 2025"
author: "John Kathena, Jim Ianelli"
date: today
quarto: true
prefer-html: true
vignette: >
  %\VignetteIndexEntry{Model update}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
lightbox: true
format:
  pdf:
        toc: true
        toc-title: 'Contents'
  html:
        code-summary: "Show the code"
        embed-resources: true
        toc: true
        toc-title: 'Contents'
        number-sections: false
        code-fold: true
editor: 
  markdown: 
    wrap: 72
project:
  render: ['*.qmd']
       # - name: Set up Quarto
       # uses: quarto-dev/quarto-actions/setup@v2
editor_options: 
  chunk_output_type: console
---

```{r setup1}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x <- knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x <- strwrap(x, width = n)
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(collapse = TRUE, comment = "  ", fig.align = "center", cache = FALSE, tidy.opts = list(width.cutoff = 80), tidy = TRUE)
knitr::opts_knit$set(root.dir = here::here())
# knitr::opts_chunk$set(warning=F, message=F, echo=F, results=F,fig.width=6, fig.height=5)
```

# Assessment model runs

The original base-case model was evaluated for a number of features and
extensions. These included focus on what data components were fit well
and how improvements in consistency can be made. For the latter part, we
found that the fits to the index and CPUE data were particularly poor
and could be improved. This document is an update from the  base-case model presented
in 2024. 

### Model descriptions

The following table was developed based on testing the model with
different assumptions and data sources. Key differences from the 2023
assessment configuration was the assumption that model estimation of
variance terms was appropriate. This feature resulted in unacceptable
residual patterns and essentially a complete down weighting of the index
data. We used the assumed variance terms (CVs) for the indices in all of
the following model configurations:

| Model | Description |
|-------------------------|-----------------------------------------------|
| Pre 2024 model configuration | As specified in past assessments, estimated steepness and all variance terms |
| Base case (m0) | Model with survey "minus group" to be ages 0, and 1 instead of 0, 1, and 2 as done in the past, steepness fixed at 0.7, q estimated, and time-varying fishery asymptotic selectivity specified. |
| m25 | As base case but updated through 2025 . |
| m1 | As 2024 base case but with survey catchability fixed at 1.0 |
| m2 | As 2024 base case but with survey catchability fixed at 0.5 |
| m3 | As 2024 base case but with natural mortality estimated |
| m4 | As 2024 base case but with fishery selectivity allowed to be dome-shaped |
| m5 | As 2024 base case but with stock-recruit steepness fixed at 0.5 |
| m6 | As 2024 base case but with stock-recruit steepness fixed at 0.9 |

<!-- | m7            | As base case but with early indices dropped (pre 1990)              | -->

```{r startup}
#| output: false

library(NamibianHake)
library(flextable)
library(here)
library(tidyverse)
library(ggridges)
theme_set(ggthemes::theme_few())
library(xtable)
library(kableExtra)
set_flextable_defaults(
  digits = 3,
  decimal.mark = ".",
  big.mark = ",",
  na_str = "<na>"
)
```

```{r getmods}
#| output: false
mod_ref <- c("old_bc", "m25", "m0", "m1", "m2", "m3", "m4", "m5", "m6")
mod_dir <- c("old_bc", "m25", "m0", "m1", "m2", "m3", "m4", "m5", "m6")
mod_label <- c("2023 base case", "2025 Update", "2024 base case", "Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6")


#---Main code that extracts all the results from the model lists above------
res <- get_results(mod_names. = mod_label, moddir = mod_dir)

modlst <- res$modlst
old_bc <- modlst[[1]]
m0 <- modlst[[3]]
m25 <- modlst[[2]]
moddiag <- res$moddiag
dfsrr <- data.frame()
for (i in 1:length(mod_ref)) {
  dfsrr <- rbind(dfsrr, data.frame(
    Model = mod_label[i],
    SSB = modlst[[i]]$SSB,
    R = modlst[[i]]$Pred_Rec
  ))
}
mods <- data.frame()
for (i in 1:length(mod_ref)) {
  mods <- rbind(mods, data.frame(moddiag[[i]], Model = names(moddiag[i])))
}
```

### Fits to index data

that the variance terms (CVs) were estimated. A more standard approach,
where CVs are specified based on the data (e.g., from design-based
sampling theory), was used and this performed better (@fig-fitidx1,
@fig-fitcpue3). In the previous assessment, the fit tot he early period
of CPUE data was better, but in this case the CV was estimated to be
about 10%, and extremely low value for this type of data
(@fig-fitcpue1).

Similarly, we re-evaluated the ability of this model to estimate the
stock recruitment productivity parameter (steepness). It is extremely
rare that sufficient data are available to freely estimate this, even
with extensive high-quality index data. Therefore we evaluated what
assumptions were taken elsewhere (for this and related species, e.g.,
South African hake) and ran the models with steepness fixed at 0.7.

```{r fitsurv}
#| echo: false
#| fig.cap: "Base case model fits to main survey data compared to the previous assessment."
#| label: fig-fitidx1


dfidx <- rbind(
  # data.frame(Year = 1964:2023, Obs = old_bc$Obs_Survey_1, predicted = old_bc$Pre_Survey_1, Model = "Base case 2023"),
  data.frame(Year = 1964:2024, Obs = m0$Obs_Survey_1, predicted = m0$Pre_Survey_1, Model = "Base case 2024"),
  data.frame(Year = 1964:2025, Obs = m25$Obs_Survey_1, predicted = m25$Pre_Survey_1, Model = "Base case 2025")
)
dfidx |>
  filter(Obs > -0) |>
  ggplot(aes(x = Year, y = Obs, color = Model)) +
  geom_point(color = "black") +
  ylim(0, NA) +
  geom_line(aes(y = predicted)) +
  geom_point(aes(y = predicted))
ggsave(here("mods", "figs", "fitidx1.png"), width = 6, height = 5)
```

```{r fitcpue3}
#| echo: false
#| fig.cap: "Base case model fits to the CPUE index 3 data compared to the previous assessment."
#| label: fig-fitcpue3

dfcpue <- rbind(
  data.frame(Year = 1964:2024, Obs = m0$Obs_CPUE_3, predicted = m0$e_CPUE_3, Model = "Base case 2024"),
  data.frame(Year = 1964:2025, Obs = m25$Obs_CPUE_3, predicted = m25$e_CPUE_3, Model = "Base case 2025")
)

dfcpue |>
  filter(Obs > 0) |>
  ggplot(aes(x = Year, y = Obs, color = Model)) +
  geom_point(color = "black") +
  geom_line(aes(y = predicted)) +
  geom_point(aes(y = predicted)) +
  ggtitle("CPUE 3") +
  ylim(0, NA)
ggsave(here("mods", "figs", "cpue3.png"), width = 6, height = 5)
```

```{r fitcpue1}
#| fig.cap: "Base case model fits to the CPUE index 1 data compared to the previous assessment."
#| label: fig-fitcpue1
dfcpue <- rbind(
  data.frame(Year = 1964:2024, Obs = m0$Obs_CPUE_1, predicted = m0$e_CPUE_1, Model = "Base case 2024"),
  data.frame(Year = 1964:2025, Obs = m25$Obs_CPUE_1, predicted = m25$e_CPUE_1, Model = "Base case 2025")
)
dfcpue |>
  filter(Obs > 0) |>
  ggplot(aes(x = Year, y = Obs, color = Model)) +
  geom_point(color = "black") +
  geom_line(aes(y = predicted)) +
  geom_point(aes(y = predicted)) +
  ggtitle("CPUE 1") +
  ylim(0, NA)
ggsave(here("mods", "figs", "cpue1.png"), width = 6, height = 5)
```

```{r fitcpue3_all}
#| fig.cap: "Base case model fits to the CPUE index 3 data compared to 2025 and also models 1 and 2."
#| label: fig-fitcpue1_all
dfcpue <- rbind(
  data.frame(Year = 1964:2024, Obs = m0$Obs_CPUE_3, predicted = m0$e_CPUE_3, Model = "Base"),
  data.frame(Year = 1964:2025, Obs = m25$Obs_CPUE_3, predicted = m25$e_CPUE_3, Model = "Base updated 2025"),
  data.frame(Year = 1964:2024, Obs = modlst[[3]]$Obs_CPUE_3, predicted = modlst[[3]]$e_CPUE_3, Model = "Model 1, q=1.0"),
  data.frame(Year = 1964:2024, Obs = modlst[[4]]$Obs_CPUE_3, predicted = modlst[[4]]$e_CPUE_3, Model = "Model 2, q=0.5")
)
dfcpue |>
  filter(Obs > 0) |>
  ggplot(aes(x = Year, y = Obs, color = Model)) +
  geom_point(color = "black") +
  geom_line(aes(y = predicted)) +
  geom_point(aes(y = predicted)) +
  ggtitle("CPUE 1") +
  ylim(0, NA)
ggsave(here("mods", "figs", "cpue1_all.png"), width = 6, height = 5)
```

```{r fitsurv_all}
#| echo: false
#| fig.cap: "Base case model fits to main survey data compared to models 1 and 2."
#| label: fig-fitidx1_all

dfidx <- rbind(
  data.frame(Year = 1964:2024, Obs = m0$Obs_Survey_1, predicted = m0$Pre_Survey_1, Model = "Base 24"),
  data.frame(Year = 1964:2025, Obs = m25$Obs_Survey_1, predicted = m25$Pre_Survey_1, Model = "Base 2025 "),
  data.frame(Year = 1964:2024, Obs = m0$Obs_Survey_1, predicted = modlst[[3]]$Pre_Survey_1, Model = "Model 1, q=1.0"),
  data.frame(Year = 1964:2024, Obs = m0$Obs_Survey_1, predicted = modlst[[4]]$Pre_Survey_1, Model = "Model 2, q=0.5")
)
dfidx |>
  filter(Obs > -0) |>
  ggplot(aes(x = Year, y = Obs, color = Model)) +
  geom_point(color = "black") +
  ylim(0, NA) +
  geom_line(aes(y = predicted)) +
  geom_point(aes(y = predicted))
ggsave(here("mods", "figs", "fitidx1_all.png"), width = 6, height = 5)
```

### Age composition fits

The age composition fits for the base case model are shown inf
@fig-agecomps1 and @fig-agecomps2. The base case model uses a 'minus
group' equal to '1' for the survey data and for the fishery it was set
to 2 (as was done previously).

```{r agecomps1}
#| fig.cap: "Base case model fits to the fishery age composition data. Note that the
#| base case model uses a 'minus group' equal to '2' for the fishery data."
#| label: fig-agecomps1
PlotAgeFit(x = m25, title = "Base case 2025", type = "fishery", fage = 2, lage = 7) + ggthemes::theme_few(base_size = 10)
ggsave(here("mods", "figs", "Age_comp_fish.png"), width = 9, height = 8)
```

```{r agecomps2}
#| fig.cap: "Base case model fits to the survey age composition data. Note that the
#| base case model uses a 'minus group' equal to '1' for the survey data."
#| label: fig-agecomps2
PlotAgeFit(x = m25, title = "Base case 2025", type = "survey1", fage = 1, lage = 7) + ggthemes::theme_few(base_size = 10)
ggsave(here("mods", "figs", "Age_comp_surv.png"), width = 9, height = 8)
```

### Selectivity

```{r selex}
#| echo: false
#| fig.cap: "Selectivity estimates for the updated to 2025 base-case model run."
#| fig.width: 4
#| fig.height: 9

M <- data.frame(Year = 1964:2024, Sel = m25$S[1:61, ], Model = "h=0.7, base case")
p1 <- plot_sel()
# p1
ggsave(here("mods", "figs", "selectivity.png"), width = 5, height = 9)
```

### Spawning biomass trends

@fig-ssb shows the SSB estimates for the base case model compared to the
previous assessment. The horizontal lines correspond to the Bmsy values
for the separate models.

```{r SSB}
#| fig.cap: "Base case model showing the SSB estimates compared to the previous assessment. The horizontal
#| lines correspond to the Bmsy values for the separate models."
#| label: fig-ssb
dftmp <- data.frame(Model = mod_label[1], Bmsy = (modlst[[1]]$Bmsy))
dftmp <- rbind(dftmp, data.frame(Model = mod_label[2], Bmsy = (modlst[[2]]$Bmsy)))
mods |>
  filter(Model %in% mod_label[2:3], Year > 1960, Variable == "SSB") |>
  ggplot(aes(x = Year, y = value, ymin = ymin, ymax = ymax, type = Model, fill = Model)) +
  geom_ribbon(alpha = .24) +
  ggthemes::theme_few() +
  coord_cartesian(ylim = c(0, 6e3)) +
  geom_line(aes(color = Model)) +
  geom_hline(yintercept = dftmp$Bmsy[1], color = 2) +
  geom_hline(yintercept = dftmp$Bmsy[2], color = 3) +
  geom_point(aes(color = Model, shape = Model), size = 1.) +
  ylab("SSB") +
  xlab("Year")
ggsave(here("mods", "figs", "ssb1.png"), width = 6, height = 5)
```

Another way to evaluate the relative trends in spawning biomass is to
examine the so-called "depletion" levels. This is the ratio of the
current SSB to the theoretical unfished value. The results are shown in
@fig-depl for the base case model and in @fig-deplalt for the
alternative models.

```{r depl}
#| fig.cap: "Base case model showing the relative SSB estimates compared to the previous assessment."
#| label: fig-depl
mods |>
  filter(Model %in% mod_label[3:2], Year > 1960, Variable == "Depletion") |>
  ggplot(aes(x = Year, y = value, ymin = ymin, ymax = ymax, type = Model, fill = Model)) +
  geom_ribbon(alpha = .24) +
  ggthemes::theme_few() +
  geom_line(aes(color = Model)) +
  geom_point(aes(color = Model, shape = Model), size = 1.) +
  ylab("Relative SSB") +
  xlab("Year")
ggsave(here("mods", "figs", "depl1.png"), width = 6, height = 5)
```

```{r deplalt}
#| fig.cap: "Alternative model results showing the relative SSB estimates."
#| label: fig-deplalt
mods |>
  filter(Model %in% mod_label[2:9], Year > 1960, Variable == "Depletion") |>
  ggplot(aes(x = Year, y = value, ymin = ymin, ymax = ymax, type = Model, fill = Model)) +
  geom_ribbon(alpha = .24) +
  ggthemes::theme_few() +
  geom_line(aes(color = Model)) +
  geom_point(aes(color = Model, shape = Model), size = 1.) +
  ylab("Relative SSB") +
  xlab("Year")
ggsave(here("mods", "figs", "depl2.png"), width = 6, height = 5)
```

### Recruitment estimates

The recruitment results are consistent with those seen for the spawning
biomass. The recent models shows a decline in recruitment in general, and that
given the new data, the estimates for the 2024 year-class dropped from the mean
value assumed from the 2024 assessment (@fig-rec). Compared to
the base case model, the alternative 2024 models show a range of recruitment
estimates (@fig-recalt).

```{r rec}
#| fig.cap: "Alternative model results on recruitment estimates."
#| label: fig-rec
mods |>
  filter(Model %in% mod_label[3:2], Year > 2010, Variable == "R") |>
  ggplot(aes(x = Year, y = value, ymin = ymin, ymax = ymax, type = Model, fill = Model)) +
  geom_errorbar(width = 0.95, position = "dodge", alpha = .3) +
  geom_bar(width = 0.95, stat = "Identity", position = "dodge") +
  ggthemes::theme_few() +
  ylab("Recruitment age 0") +
  xlab("Year")
ggsave(here("mods", "figs", "rec.png"), width = 6, height = 5)
```

```{r recalt}
#| fig.cap: "Alternative model results on recruitment estimates."
#| label: fig-recalt
mods |>
  filter(Model %in% mod_label[2:9], Year > 2010, Variable == "R") |>
  ggplot(aes(x = Year, y = value, ymin = ymin, ymax = ymax, type = Model, fill = Model)) +
  geom_errorbar(width = 0.95, position = "dodge", alpha = .3) +
  geom_bar(width = 0.95, stat = "Identity", position = "dodge") +
  ggthemes::theme_few() +
  ylab("Recruitment age 0") +
  xlab("Year")
ggsave(here("mods", "figs", "recalt.png"), width = 6, height = 5)
```

### Stock recruitment relationships

For each of the models there were some specified and estimated
differences in the stock-recruitment relationships (@fig-srrplots). When
overlain with the recruitment estimates, these relationships appear to
be relatively poorly defined (@fig-srr).

```{r srrplots}
#| fig.cap: "Stock-recruitment curves noted for the previous base-case and
#| the updated model alternatives."
#| label: fig-srrplots
p1 <- dfsrr |> ggplot(aes(x = SSB, y = R, color = Model, shape = Model)) +
  geom_point() +
  geom_line() +
  xlim(c(0, 4000))
p1
ggsave(here("mods", "figs", "srr_curves.png"), width = 6, height = 5)
```

```{r srr}
#| label: fig-srr
#| fig.cap: "Stock-recruitment curves and year-class estimates for alternative models"
#| fig.width: 9
#| fig.height: 7
#|
mods |>
  filter(Model %in% mod_label[2:9], Variable == "R" | Variable == "SSB") |>
  select(c(1:3, 6)) |>
  pivot_wider(names_from = Variable, values_from = value) |>
  ggplot(aes(x = SSB, y = R, label = Year, color = Model, fill = Model)) +
  geom_text(alpha = .7, size = 3) +
  ggthemes::theme_few() +
  facet_wrap(. ~ Model) +
  geom_line(
    data = dfsrr |> filter(Model %in% mod_label[2:9]),
    aes(y = R, x = SSB, color = Model), inherit.aes = FALSE
  )
ggsave(here("mods", "figs", "srr_yrs.png"), width = 6, height = 5)
```

To show the history relative to the replacement yield, we can plot the
SSB/Bmsy and the catch/replacement yield (@fig-kobe). This shows a
significant difference between the previous assessment and that proposed
as the base-case for this year. The alternative models show a range of
results, as shown in @fig-kobe2.

```{r kobe}
#| fig.cap: "Base case model showing the relative SSB estimates compared to the 2024 assessment."
#| label: fig-kobe
tmp <- mods |>
  select(Year, Model, Variable, value) |>
  filter(Model %in% mod_label[2:3], Year > 1980, Variable %in% c("Catch_RY", "B_Bmsy")) |>
  pivot_wider(names_from = Variable, values_from = value) |>
  arrange(Model, Year) #|>
tmp |> ggplot(aes(x = B_Bmsy, label = Year, y = Catch_RY, shape = Model, color = Model, fill = Model)) +
  geom_path() +
  ggthemes::theme_few() +
  geom_point() +
  geom_text(alpha = .5) +
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) +
  xlab("SSB/Bmsy") +
  ylab("Catch / replacement Yield")
ggsave(here("mods", "figs", "kobe1.png"), width = 6, height = 5)
```

```{r kobe2}
#| fig.cap: "Base case model showing the relative SSB estimates compared to the
#| model alternatives assessment."
#| label: fig-kobe2
tmp <- mods |>
  select(Year, Model, Variable, value) |>
  filter(Model %in% mod_label[2:3], Year > 1980, Variable %in% c("Catch_RY", "Depletion")) |>
  pivot_wider(names_from = Variable, values_from = value)
tmp <- mods |>
  select(Year, Model, Variable, value) |>
  filter(Model %in% mod_label[c(2:3, 7:8)], Year > 1980, Variable %in% c("Catch_RY", "B_Bmsy")) |>
  pivot_wider(names_from = Variable, values_from = value) |>
  arrange(Model, Year) #|>
tmp |> ggplot(aes(x = B_Bmsy, label = Year, y = Catch_RY, shape = Model, color = Model, fill = Model)) +
  geom_path() +
  ggthemes::theme_few() +
  geom_point() +
  geom_text(alpha = .5) +
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) +
  xlab("SSB/Bmsy") +
  ylab("Catch / replacement Yield")
ggsave(here("mods", "figs", "kobe2.png"), width = 6, height = 5)
```

### Stock status comparisons

The @tbl-quant show the results of the base case model compared to the
alternative models for a number of key statistics. Among these, the best
fitting models were the base-case and Model 5. However, model 5 fits
indicated that the basis for fitting the data was virtually identical
and that the slight benefit based on the AIC arises from the
stock-recruitment residuals. For advice and consistency with
productivity assumptions (i.e., the specification of the steepness
parameter fixed at 0.7 as done for South Africa) we selected the
base-case scenario.

```{r mgt_quant, results = "asis"}
#| echo: true
#| results: asis
#| label: tbl-quant
#| tbl-cap: "Model fits to data components"
#|
dftmp <- NULL
mod_scen <- c(2:3, 5:8)
filler <- " "
names(filler) <- "-ln(Likelihood)"
for (ii in mod_scen) {
  x <- modlst[[ii]]
  nll <- round(x$ObjFun, 0)
  names(nll) <- "Overall"
  CPUE <- round(x$CPUE_Like, 0)
  names(CPUE) <- "CPUE"
  surv <- round(x$Survey_Like, 0)
  names(surv) <- "Survey"
  caa <- round(x$CAA_Likelihood, 0)
  names(caa) <- "Commercial CAA"
  caas <- round(x$CAAS_Likelihood, 0)
  names(caas) <- "Survey CAA"
  oneyrold <- round(x$Oneyearold_Likelihood, 0)
  names(oneyrold) <- "One yr-old biomass"
  rec <- round(x$RecRes_Likelihood, 0)
  names(rec) <- "Rec. resids."
  datalike <- nll - rec
  names(datalike) <- "Data likelihood sub-total"
  NumPars <- round(x$Npars, 0)
  names(NumPars) <- "Number parameters"
  AIC <- round(x$Akaike, 1)
  names(AIC) <- "AIC"
  v <- c(filler, nll, CPUE, surv, caa, caas, oneyrold, datalike, rec, NumPars, AIC)
  dftmp <- cbind(dftmp, v)
}
dftmp <- data.frame(rownames(dftmp), dftmp, row.names = NULL)
names(dftmp) <- c("Component", mod_label[mod_scen])
tabcap <- "Model fits to data components"
flextable(dftmp) |>
  set_caption(caption = tabcap) |>
  colformat_double() |>
  autofit()
```

```{r status}
#| label: tbl-stockstatus
#| tbl-cap: "Namibian hake stock estimates by model alternative."
dftmp <- NULL
mod_scen <- c(2:3)
for (ii in mod_scen) {
  x <- modlst[[ii]]
  Ksp <- round(x$KspSTD, 0)
  names(Ksp) <- "Unfished spawning biomass"
  Kexp <- round(x$KexpSTD, 0)
  names(Kexp) <- "Unfished expl. biomass"
  steepness <- round(x$Steep, 3)
  names(steepness) <- "SRR steepness"
  Cur_B <- round(x$Bstd[length(x$Bstd)], 3)
  names(Cur_B) <- "2024 SSB"
  Bmsy <- round(x$Spmsy, 0)
  names(Bmsy) <- "SSB_msy"
  Cur_B0 <- round(x$Cur_B0, 3)
  names(Cur_B0) <- "Current SSB over unfished"
  Cur_Bmsy <- round(x$Cur_Bmsy, 3)
  names(Cur_Bmsy) <- paste0("Current SSB over Bmsy")
  MSY <- round(x$MSY, 0)
  names(MSY) <- "MSY"
  ry <- round(x$aveRY_last5, 0)
  names(ry) <- paste0("recent 5-yr average replacement yield")
  ry_cur <- round(x$aveRY_last5 * x$Cur_Bmsy, 3)
  names(ry_cur) <- "Recent RY x current SSB /Bmsy"
  v <- c(Ksp, Kexp, steepness, Cur_B, Bmsy, Cur_B0, Cur_Bmsy, MSY, ry, ry_cur)
  dftmp <- cbind(dftmp, v)
}
dftmp <- data.frame(rownames(dftmp), dftmp, row.names = NULL)
names(dftmp) <- c("Statistic", mod_label[mod_scen])
tabcap <- "Selected management measures from alternative models. "
flextable(dftmp) |>
  set_caption(caption = tabcap) |>
  colformat_double() |>
  autofit()
# align=paste0("lll",strrep("r",length(mod_scen+1))))
# kable(tab, caption.placement = "top", include.rownames = FALSE, sanitize.text.function = function(x){x})
# print(tab)
```

A comparison among the models for stock status and reference points shows that
the 2025 update is similar to the 2024 base case, but that the alternative models
show a range of results (@fig-stockstatus). 

```{r stockstatus}
#| echo: false
#| label: fig-stockstatus
#| fig.cap: "Some reference point comparisons  among model runs.
#| aveRY_90 is the average replacement yeild since 1990, aveRY is the average replacement yield over the most recent 5 years before the current year, Cur_90 is
#| the current (terminal year) SSB over the estimate from 1990, Cur_B0 is over the unfished estimate, and Cur_Bmsy is the ratio of current SSB over the Bmsy estimate. "
#| fig.width: 8
#| fig.height: 5

#--Plot the reference points-------------------
mods |>
  filter(Model != "2023 base case", Variable %in% c("aveRY_90", "aveRY_last5", "Cur_90", "Cur_B0", "Cur_Bmsy")) |>
  ggplot(aes(x = Model, y = value, ymin = ymin, ymax = ymax, color = Model, fill = Model)) +
  geom_point(size = 3) +
  ggthemes::theme_few() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_errorbar() +
  ylab("Value") +
  xlab("") +
  facet_wrap(. ~ Variable, scales = "free")
ggsave(here("mods", "figs", "refpoints.png"), width = 6, height = 5)
```

# Control rule application

## Modeling Namibian hake survey data by species

Fisheries stock assessments require data that are reliably collected and
compiled. Secondarily, assessment models should be configured to match
the assumptions associated with the observed data. To account for survey
trends between the two species of hake we applied the estimated
observation errors to a simple state-space random walk model. This
approach has a number of options for how process-errors can be specified
and estimated. The observation model applies the observation-error
variances $(\sigma_{j,t}^2)$ for the $j^{th}$ species in year
$t (x_{j,t} )$. The indices are fit to latent state variables, e.g., the
underlying population trend $ln(\hat{Z}_{j,t})$ as follows:

$ln(Z_{j,t} ) = ln(\hat{Z}_{j,t} )+\epsilon_{j,t}$ where
$\epsilon_{j,t}∼N(0,\sigma_{j,t}^2 )$

and the state equation and associated process error variance
$\sigma_{PE}^2$ is defined as

$ln(\hat{Z}_{j,t+1} = ln(\hat{Z}_{j,t+}  )+\eta_{j,t},$ where
$\eta_{j,t}∼N(0,\tau_j^2 ).$

The process error variances $\tau_j^2$ (which may or may not vary across
indices) are fixed effect parameters and the unobserved species combined
population $ln(Z_{j,t} )$ is estimated as a series of random effects.
The model is fit using maximum likelihood estimation in TMB using the R
package “rema” (Sullivan 2022, [rema
library](https://afsc-assessments.github.io/rema)). The survey data for
each species was used with CVs applied for observation error
specifications. The values for $\tau_j^2$ were tested for each species
and found to be similar so they were set to the same values.

The above analysis provides a summary of the model runs and the design
of a control rule that accounts for the signals in the data on the
different species.

## Application to the Namibian hake stocks

The control rule was first applied to the Namibian hake stocks using the
survey data and the relative proportions of the two species. However, we
wish to have a more robust control rule that accounts for the signals in
the data on *M. paradoxus* biomass and have that be independent of *M.
capensis* (e.g., @fig-relmean). For that case, we applied the survey
smoothing model to *paradoxus* alone. The next step was to compute mean
biomass over the period 1990-2024, and evaluate the adustment for
different levels of $\gamma$. The historical adjustments based on this
aspect of the MP is shown in @fig-adjust.

```{r relmean}
#| echo: false
#| fig.width: 6
#| fig.height: 5
#| message: false
#| warning: false
#| results: asis
#| label: fig-relmean
#| fig.cap: "Historical biomass estimates for M. paradoxus and the mean biomass over the period 1990-2025."

biom_out <- get_relative_biomass()
# tail(biom_out$total_predicted_biomass)

p1 <- biom_out$biomass_by_strata |>
  filter(strata == "paradoxus") |>
  ggplot(aes(x = year, ymin = pred_lci, ymax = pred_uci, fill = strata, y = obs)) +
  geom_ribbon(alpha = .6) +
  scale_y_continuous(labels = scales::comma, expand = c(0, 0), limits = c(0, NA)) +
  geom_errorbar(aes(x = year, ymin = obs_lci, ymax = obs_uci), width = .2, alpha = .4) +
  geom_line(aes(x = year, y = pred)) +
  ylab("Biomass (kt)") +
  ggthemes::theme_few() +
  geom_point(alpha = .7) +
  theme(legend.position = "none")
mean_mp <- exp(mean(p1$data$log_pred))
p1 + geom_hline(yintercept = mean_mp, linetype = "dashed", color = "blue")
ggsave(here("mods", "figs", "biom_paradoxus.png"), width = 6, height = 5)
```

```{r relmean}
#| echo: false
#| label: fig-adjust
#| fig.cap: "Historical adjustments for different reactivities to *M. paradoxus* biomass."
#---Plot the adjustment factors for different reactivities to M. paradoxus biomass----
dftmp <- p1$data |>
  select(Year = year, Index = pred) |>
  mutate(
    "Gamma = 0.1" = if_else(Index / mean_mp > 1, 1, Index / mean_mp)^0.1,
    "Gamma = 0.5" = if_else(Index / mean_mp > 1, 1, (Index / mean_mp)^.5),
    "Gamma = 0.25" = if_else(Index / mean_mp > 1, 1, (Index / mean_mp)^.25)
  ) |>
  pivot_longer(cols = -Year, names_to = "Reactivity", values_to = "Index")
dftmp |>
  filter(Reactivity != "Index") |>
  ggplot(aes(x = Year, y = Index, color = Reactivity)) +
  geom_line() +
  geom_point() +
  ggtitle("Historical adjustments for different reactivities to *M. paradoxus* biomass") +
  ggthemes::theme_few() +
  ylab("TAC adjustment based on *M. paradoxus* biomass")
ggsave(here("mods", "figs", "adj_paradoxus.png"), width = 6, height = 5)
cur_mp <- dftmp |>
  filter(Year == 2025, Reactivity == "Gamma = 0.25") |>
  pull(Index)
cur_mp <- round(cur_mp, 3)
```

## Control rule developments

The situation for developing a two-species control rule where catches
between species and the trend in overall biomass for both species is
combined is challenging. For management purposes, the goal is to avoid
incidental takes in the proportion of one species that exceeds the
historical levels of depletion for either stock. Fortunately, survey
data are available that can be used to distinguish trends in the
relative biomass for both stocks. The design of the triggered control
rule therefore must consider patterns in the relative biomass from the
survey data, the absolute biomass of the combined catch and biomass as
modeled from the combined-stocks assessment. This provides a pragmatic
approach using available data.

The steps in the control rule would be first to run a simple model that
projected the survey biomass and relative proportion of *M. paradoxus*.
Then, given the mean proportion over the period, compute the adjustment
needed to the overarching control rule for the management procedure. For
example, the historical range based on the survey has been between about
one third of the mean value (in the earliest part of the period) to
about 70% above the mean proportion. This range (especially the lower
value) was used as a semi-empirical way to develop a minimum stock size
threshold as part of the control rule. That is, when the stock of *M.
paradoxus* drops below 30% of the mean proportion of the combined stocks
estimate, the TAC recommendation for the combined stock would be zero.

$$
TAC_y = \begin{cases} 
0 & \text{if } p_y < 0.3\bar{p} \\
\delta \left(\frac{1}{5}\sum_{i=0}^{4} RY_{y-i}\right) \times f(B_y, \lambda) \times \min\left(1, \frac{\tilde{p}_y}{\bar{p}}\right)^\gamma & \text{otherwise}
\end{cases}
$$

where:
$$
f(B_y, \lambda) = \begin{cases}
\left(\frac{B_y/B_{MSY}}{0.5}\right)^\lambda & \text{if } B_y/B_{MSY} < 0.5 \\
1 & \text{otherwise}
\end{cases}
$$


where the rebuilding factor ($\delta$) is set to 0.8 when the spawning
biomass is below $B_{MSY}$ and 1.0 when above. In words, the TAC in year
y is equal to the catch under the current control rule times the ratio
of the spawning biomass relative to $B_{MSY}$ (or proxy) and the
externally estimated proportion of *M. paradoxus* from survey data. The
second term relates $B_{MSY}$ and is intended to take fast action (for
$\lambda >1.0$) when the biomass falls below 0.5 of that value (a
standard in many places to define "overfished"). 
The third term on the right hand side reflects the impact of the *M. paradoxus* proportion projected from a survey smoother (described below) and adjusts the TAC advice downwards when the projected $\tilde{p}_y$ drops below the mean value.

The values of $\gamma, \lambda$ were evaluated and are shown in
@tbl-controlrule. We note that the specification of a survey linkage by
the individual species provides an appropriate adjustment that reduces
the exploitation rate and prevents potential for "the point of
recruitment impairment" (PRI).

For the control rule as specified, the reactivity of the TAC advice to
changes in either B/Bmsy or the biomass of *M. paradoxus* biomass can be
adjusted. These are shown in @fig-gamma. For the purposes of this
analysis, we set $\gamma = .25$. 
For $\lambda$, most models evaluated were above 0.5 of $B_{MSY}$ so $f(B_y, \lambda) = 1$ and there was no added adjustment beyond $\delta=0.8$. So, following the TAC as specified, the base-case model was below $B_{MSY}$ so $\delta=0.8$ with the average replacement yield over the last 5 years is recommended, with the adjustment based on the relative proportion of *M. paradoxus* and the long-term mean proportion of the species.
The TAC advice for the combined stocks is then given in
@tbl-controlrule.

```{r controlrule}
#| label: tbl-controlrule
#| tbl-cap: "Namibian hake specification of different options for TAC considerations"
#|
dftmp <- NULL
mod_scen <- c(2:3)
ii <- 2
names(cur_mp) <- "Ratio of paradoxus to mean "
for (ii in mod_scen) {
  x <- modlst[[ii]]
  Cur_Bmsy <- round(x$Cur_Bmsy, 3)
  names(Cur_Bmsy) <- paste0("Current B over Bmsy")
  ry <- round(x$aveRY_last5, 0)
  names(ry) <- paste0("recent 5-yr avg repl. yield")

  ry_cur <- round(x$aveRY_last5 * min(1, x$Cur_Bmsy), 3)
  # names(ry_cur) <- ("Recent RY x current B/Bmsy")

  lambda <- 1.5
  gamma <- 0.25
  if (Cur_Bmsy > 0.5) adj <- 1 else adj <- Cur_Bmsy^lambda / 0.5^lambda
  tac1 <- round(0.8 * ry * adj * min(1, cur_mp)^gamma, 2)
  names(tac1) <- paste0("opt 1, lambda=", lambda, " gamma=", gamma)

  lambda <- 2
  gamma <- 0.25
  if (Cur_Bmsy > .5) adj <- 1 else adj <- Cur_Bmsy^lambda / 0.5^lambda
  tac2 <- round(0.8 * ry * adj * min(1, cur_mp)^gamma, 2)
  names(tac2) <- paste0("opt 2, lambda=", lambda, " gamma=", gamma)

  lambda <- 2
  gamma <- 0.1
  if (Cur_Bmsy > .5) adj <- 1 else adj <- Cur_Bmsy^lambda / 0.5^lambda
  adj
  Cur_Bmsy
  tac3 <- round(0.8 * ry * adj * min(1, cur_mp)^gamma, 2)
  names(tac3) <- paste0("opt 3, lambda=", lambda, " gamma=", gamma)

  lambda <- 2
  gamma <- 1.0
  if (Cur_Bmsy > .5) adj <- 1 else adj <- Cur_Bmsy^lambda / 0.5^lambda
  adj
  Cur_Bmsy
  tac4 <- round(0.8 * ry * adj * min(1, cur_mp)^gamma, 2)
  names(tac4) <- paste0("opt 4, lambda=", lambda, " gamma=", gamma)

  filler <- ""
  names(filler) <- "TAC"

  v <- c(ry, Cur_Bmsy, cur_mp, filler, tac1, tac2, tac3, tac4)
  dftmp <- cbind(dftmp, v)
}
dftmp <- data.frame(rownames(dftmp), dftmp, row.names = NULL)
names(dftmp) <- c("Statistic", mod_label[mod_scen])
tabcap <- "Selected management measures from alternative models. "
flextable(dftmp) |>
  set_caption(caption = tabcap) |>
  colformat_double() |>
  hline(i = 3) |>
  autofit()
```

```{r gamma}
#| label: fig-gamma
#| fig.cap: "TAC adjustments for different  levels of current biomass of M. paradoxus biomass."
#--- Plot the adjustment factors for different reactivities to M. paradoxus biomass----
df01 <- tibble(
  relbiom = 1:20 / 20,
  "Gamma = 0.1" = (1:20 / 20)^0.1,
  "Gamma = 0.5" = (1:20 / 20)^.5,
  "Gamma = 0.25" = (1:20 / 20)^.25,
) |> pivot_longer(cols = -relbiom, names_to = "Reactivity", values_to = "adjustment")
df01 |>
  ggplot(aes(y = adjustment, x = relbiom, color = Reactivity)) +
  geom_line() +
  xlab("Relative M. paradoxus biomass") +
  ylab("TAC adjustment") +
  ggthemes::theme_few() #+ ylab("TAC adjustment based on M. paradoxus biomass")```{r lam_gam}
ggsave(here("mods", "figs", "gamma.png"), width = 6, height = 5)
```

```{r lambda}
#| label: fig-lambda
#| fig.cap: "TAC adjustments for different  levels of current biomass relative to Bmsy "
#---Plot the adjustment factors for different reactivities to M. paradoxus biomass----
df01 <- tibble()
df01 <- rbind(
  df01,
  tibble(relbiom = 1:100 / 100, lambda = rep(0.5, 100)),
  tibble(relbiom = 1:100 / 100, lambda = rep(1, 100)),
  tibble(relbiom = 1:100 / 100, lambda = rep(1.5, 100)),
  tibble(relbiom = 1:100 / 100, lambda = rep(2, 100))
) |> mutate(adjustment = if_else(relbiom < 0.5, relbiom^lambda / 0.5^lambda, 1), lambda = as.factor(lambda))
df01 |> ggplot(aes(y = adjustment, x = relbiom, color = lambda)) +
  geom_line() +
  xlab("Spawning biomass relative Bmsy") +
  ylab("TAC adjustment") +
  ggthemes::theme_few() #+ ylab("TAC adjustment based on M. paradoxus biomass")
ggsave(here("mods", "figs", "lambda.png"), width = 6, height = 5)
```

# Projections

A set of simple projections were established from the base case model.
These were based on the 2024 catch estimates (lower and higher values)
for contrast relative to depletion @fig-proj.

```{r proj }
#| label: fig-proj
#| echo: false
#| warnings: false
#| messages: false
#| fig.cap: "40-year projections at different levels of catch related to the catch in 2025 (e.g., 0.5*C_25 is constant catch at 50% of the catch estimated for 2025). The scenario labeled 'Replacement yld' is just the catch set to the replacement yield. The cases with '10yr' represent catch scenarios where the catch is increased (and decreased) by 10% of the 2025 catch and thereafter set equal to the 2025 catch."
prj <- readr::read_csv(
  "mods/m25/main.out",
  col_names = FALSE,
  show_col_types = FALSE, # <-- suppress column spec
  progress = FALSE
)
names(prj) <- c("Scenario", "Year", "Var", "Value")
scen <- c("0.5*C_25", "0.75*C_25", "0.9*C_25", "C_25", "1.1*C_25", "1.5*C_25", "2.0*C_25", "10yr up 10%", "10yr down 10%", "Replacement yld")
prj$Projection <- scen[prj$Scenario]
# glimpse(prj)Var=="RY" | Var=="SSB" |
plotly::ggplotly(
  prj |> filter(
    Var == "Catch" | Var == "Depletion",
    Scenario %in% c(1, 4:5, 8:10)
  ) |> mutate(Scenario = as.factor(Scenario)) |>
    ggplot(aes(x = Year, y = Value, color = Projection)) +
    geom_line() +
    geom_point() +
    facet_grid(Var ~ ., scales = "free") +
    ylim(c(0, NA))
)
```

<!-- # Notes -->

<!-- Throughout the weeks, we scrutinized data inputs and found a couple of -->
<!-- inconsistencies. For example, data were provided for surveys in 2019 yet -->
<!-- there were no observations in that year. Similarly, the -->
<!-- abundance-at-length data from the surveys failed to identify strong -->
<!-- periods of persistence by cohorts. -->

<!-- Should a run be done where we ignore 7-vessel CPUE data set? -->

<!-- ```{{r run_nh}} -->
<!-- #| echo: true -->

<!-- #---Run all the models---------- -->
<!-- for (i in 1:length(mod_dir)) -->
<!--   run_nh(model = mod_dir[i]) -->
<!-- get_results() -->

<!-- ``` -->
