---
title: "Namibian hake model update, 2026"
author: "John Kathena, Jim Ianelli"
date: today
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

assessment_year <- 2026L
previous_year <- assessment_year - 1L
gradient_tolerance <- 0.001

source(here::here("R", "read-admb2.R"))
source(here::here("R", "assessment-validation.R"))

library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(tidyr)

theme_set(ggthemes::theme_few())

model_spec <- tibble::tribble(
  ~directory, ~model, ~configuration,
  "m25", "2025 baseline", "Accepted base configuration through 2025",
  "m26", "2026 update", "Unchanged base configuration with data through 2026",
  "m26_survey_q_1", "q = 1.0", "Survey catchability fixed at 1.0",
  "m26_survey_q_0_5", "q = 0.5", "Survey catchability fixed at 0.5",
  "m26_estimated_natural_mortality", "Estimated M", "Natural mortality estimated",
  "m26_dome_selectivity", "Dome selectivity", "Dome-shaped fishery selectivity",
  "m26_steepness_0_5", "h = 0.5", "Stock-recruit steepness fixed at 0.5",
  "m26_steepness_0_9", "h = 0.9", "Stock-recruit steepness fixed at 0.9"
)

read_model_csv <- function(directory) {
  read_csv(
    here::here("mods", directory, "nh_out.csv"),
    show_col_types = FALSE
  )
}

model_output <- setNames(
  lapply(model_spec$directory, read_model_csv),
  model_spec$directory
)

diagnostics <- lapply(model_spec$directory, function(directory) {
  path <- here::here("mods", directory)
  tryCatch(
    {
      x <- read_admb_convergence(path)
      tibble(
        directory = directory,
        objective = x$objective,
        max_gradient = x$max_gradient,
        minimum_hessian_eigenvalue = x$minimum_hessian_eigenvalue,
        positive_definite_hessian = x$positive_definite_hessian,
        hessian_source = x$hessian_source
      )
    },
    error = function(e) {
      header <- readLines(file.path(path, "nh.par"), n = 1L)
      tibble(
        directory = directory,
        objective = as.numeric(sub(
          ".*Objective function value = ([^ ]+).*", "\\1", header
        )),
        max_gradient = as.numeric(sub(
          ".*Maximum gradient component = ([^ ]+).*", "\\1", header
        )),
        minimum_hessian_eigenvalue = NA_real_,
        positive_definite_hessian = FALSE,
        hessian_source = "Unavailable or failed"
      )
    }
  )
}) |>
  bind_rows() |>
  left_join(model_spec, by = "directory") |>
  mutate(
    converged = is.finite(objective) &
      max_gradient <= gradient_tolerance &
      positive_definite_hessian,
    status = if_else(converged, "Converged", "Failed diagnostic criteria")
  )

terminal_value <- function(directory, variable, year) {
  model_output[[directory]] |>
    filter(Variable == variable, Year == year) |>
    pull(value)
}
```

# Executive summary

## Summary of changes in assessment inputs

The accepted 2025 model configuration was extended through
`r assessment_year` without changing the base-model assumptions. The update
adds the `r assessment_year` catch, CPUE, survey biomass, and composition rows
provided in `namhakdata26.dat`. The species-specific survey series in
`CapeHakes.csv` was extended through `r assessment_year` for *Merluccius
paradoxus* and *M. capensis*.

Automated checks confirmed continuous annual ADMB matrices from 1964 through
`r assessment_year`, the expected dimensions of all weight, catch,
composition, CPUE, and survey blocks, agreement among all terminal-year
controls, and valid `54321` and `12345` end-of-file sentinels.

## Summary of results

The `r assessment_year` base model converged with an objective function of
`r round(diagnostics$objective[diagnostics$directory == "m26"], 3)`, a maximum
gradient of
`r format(diagnostics$max_gradient[diagnostics$directory == "m26"], scientific = TRUE, digits = 3)`,
and a positive-definite Hessian. Estimated terminal spawning biomass was
`r round(terminal_value("m26", "SSB", assessment_year))` and terminal spawning
biomass relative to BMSY was
`r round(terminal_value("m26", "B_Bmsy", assessment_year), 3)`.

Four of six sensitivity configurations met the same convergence criteria as
the base model. The dome-shaped selectivity case had parameters on their lower
bounds and a singular Hessian. The model with steepness fixed at 0.5 had a
maximum gradient above 0.001 after an extended optimization. Results from
those two configurations are not used for inference.

No TAC, ABC, or OFL recommendation is made in this document. Those quantities
require the agreed Namibian management procedure and management inputs.

```{r}
#| label: tbl-summary
#| tbl-cap: "Comparison of 2025 values from the previous assessment with 2026 values from the updated assessment."

bridge_variables <- c("SSB", "Depletion", "B_Bmsy", "Catch", "RY", "Catch_RY")

bridge_table <- bind_rows(
  model_output$m25 |>
    filter(Year == previous_year, Variable %in% bridge_variables) |>
    mutate(Model = "Previous assessment (2025 value)"),
  model_output$m26 |>
    filter(Year == assessment_year, Variable %in% bridge_variables) |>
    mutate(Model = "Updated assessment (2026 value)")
) |>
  select(Model, Variable, value) |>
  pivot_wider(names_from = Model, values_from = value) |>
  mutate(
    `Change (2026 minus 2025)` =
      `Updated assessment (2026 value)` -
        `Previous assessment (2025 value)`
  ) |>
  rename(Quantity = Variable)

kable(bridge_table, digits = 3)
```

# Introduction

This operational update applies the 2025 base-model structure to information
available through `r assessment_year`. Stable model methods are not repeated
in full; the emphasis is on new inputs, bridge results, model diagnostics, and
sensitivity to the principal structural assumptions considered in the
previous assessment.

# Fishery and management history

The assessment input contains annual catches from 1964 through
`r assessment_year`. Management-history text and the final
`r assessment_year` management decision should be inserted following review
by MFMR.

`TODO: add the final 2026 TAC decision and any in-season management changes.`

# Data

```{r}
#| label: tbl-input-validation
#| tbl-cap: "Automated validation of the 2026 model inputs."

input_check <- validate_nh_inputs(here::here("mods", "m26", "nh.dat"))
input_summary <- tibble(
  Check = c(
    "Annual model years",
    "Weight-at-age dimensions",
    "Fishery composition dimensions",
    "CPUE dimensions",
    "Survey dimensions",
    "Control EOF sentinel",
    "Data EOF sentinel"
  ),
  Result = c(
    paste(range(input_check$data$years), collapse = "–"),
    paste(dim(input_check$data$weight_start), collapse = " × "),
    paste(dim(input_check$data$fishery_caa), collapse = " × "),
    paste(dim(input_check$data$cpue), collapse = " × "),
    paste(dim(input_check$data$survey), collapse = " × "),
    input_check$control$eof,
    input_check$data$eof
  )
)
kable(input_summary)
```

## Survey data

```{r}
#| label: fig-species-survey
#| fig-cap: "Species-specific survey biomass observations through 2026."

cape_hakes <- read_csv(
  here::here("mods", "data", "CapeHakes.csv"),
  show_col_types = FALSE
)

ggplot(cape_hakes, aes(year, biomass, colour = strata)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = biomass / exp(sqrt(log1p(cv^2))),
      ymax = biomass * exp(sqrt(log1p(cv^2)))
    ),
    width = 0.15,
    alpha = 0.35
  ) +
  labs(x = "Year", y = "Survey biomass", colour = "Species")
```

## Other data

No additional external data series were introduced in this update. Missing
survey observations remain represented as missing or zero according to the
existing model input convention; no values were imputed to create artificial
annual continuity.

# Analytic approach

The age-structured ADMB model and 2025 base assumptions were retained. The
control file specifies the terminal year in four locations: the model terminal
year, final selectivity period, final recruitment-estimation year, and final
Ksp period. All four are set to `r assessment_year`.

The bridge comparison changes only the data and terminal year. Sensitivities
fix survey catchability at 1.0 and 0.5, estimate natural mortality, allow
dome-shaped fishery selectivity, and fix stock-recruit steepness at 0.5 and
0.9. Convergence requires a finite objective, maximum gradient no greater
than `r gradient_tolerance`, and a positive-definite ADMB Hessian.

# Results

## Bridge comparison

```{r}
#| label: fig-bridge-ssb
#| fig-cap: "Spawning biomass from the unchanged 2025 model and the 2026 data update."

bridge_series <- bind_rows(
  model_output$m25 |> mutate(Model = "2025 baseline"),
  model_output$m26 |> mutate(Model = "2026 update")
) |>
  filter(Variable == "SSB")

ggplot(bridge_series, aes(Year, value, colour = Model, fill = Model)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.15, colour = NA) +
  geom_line() +
  labs(x = "Year", y = "Spawning biomass")
```

```{r}
#| label: fig-bridge-depletion
#| fig-cap: "Estimated depletion from the 2025 baseline and 2026 update."

bridge_series |>
  select(-Variable) |>
  bind_rows(
    model_output$m25 |>
      filter(Variable == "Depletion") |>
      mutate(Model = "2025 baseline"),
    model_output$m26 |>
      filter(Variable == "Depletion") |>
      mutate(Model = "2026 update")
  ) |>
  filter(Year >= 1990, !is.na(Variable)) |>
  ggplot(aes(Year, value, colour = Model, fill = Model)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.15, colour = NA) +
  geom_line() +
  labs(x = "Year", y = "Spawning biomass / unfished spawning biomass")
```

{{< pagebreak >}}

## Model diagnostics and sensitivities

```{r}
#| label: tbl-convergence
#| tbl-cap: "Convergence diagnostics for the bridge and sensitivity models."
#| tbl-pos: H

diagnostics |>
  select(
    Model = model,
    Objective = objective,
    `Maximum gradient` = max_gradient,
    Status = status
  ) |>
  kable(digits = 5)
```

```{r}
#| label: fig-sensitivity
#| fig-cap: "Terminal spawning biomass for the converged 2026 models."

converged_directories <- diagnostics |>
  filter(converged, directory != "m25") |>
  pull(directory)

sensitivity_terminal <- lapply(converged_directories, function(directory) {
  model_output[[directory]] |>
    filter(Year == assessment_year, Variable == "SSB") |>
    mutate(directory = directory)
}) |>
  bind_rows() |>
  left_join(model_spec, by = "directory")

ggplot(
  sensitivity_terminal,
  aes(reorder(model, value), value, ymin = ymin, ymax = ymax)
) +
  geom_errorbar(width = 0.2) +
  geom_point(size = 2) +
  coord_flip() +
  labs(x = NULL, y = "Terminal spawning biomass")
```

# Projections and harvest recommendations

The ADMB model produced replacement-yield and long-term reference-point
outputs for each converged configuration. A management recommendation is not
presented because the final control-rule inputs and decision framework were
not supplied as part of this update.

`TODO: insert the agreed harvest-control-rule calculation and resulting TAC
advice after management review.`

# Risk table and recommendation

The 2026 base model passed the declared numerical diagnostics. Structural
uncertainty remains material: estimated natural mortality and fixed survey
catchability change the inferred scale of the stock, while the dome-shaped
selectivity case is not estimable under the current parameterization. The
non-converged `h = 0.5` and dome-selectivity results should not be assigned
inferential weight.

# Status determination

The base-model estimate of spawning biomass relative to BMSY is reported in
@tbl-summary. Formal legal or management status terminology should be supplied
by MFMR using the applicable Namibian framework.

# Ecosystem considerations

No new ecosystem covariates were incorporated in the assessment likelihood.
The species-specific survey series is retained for evaluating the relative
signals for *M. paradoxus* and *M. capensis*.

# Data gaps and research priorities

1. Resolve the non-identifiability of dome-shaped selectivity before using
   that alternative for management inference.
2. Profile or otherwise investigate the `h = 0.5` optimization surface.
3. Document the provenance and preprocessing of each annual input block.
4. Add retrospective analyses and likelihood profiles to the operational
   workflow.
5. Archive the final management-procedure inputs alongside the accepted
   assessment.

# Acknowledgements

The assessment update was prepared by John Kathena and Jim Ianelli with data
provided by the Namibian Ministry of Fisheries and Marine Resources.

# Auxiliary files

The assessment source, annual controls, automated validation tests, ADMB
template, and model outputs are maintained in the `NamibianHake` repository.
