---
title: "Namibian hake model update, 2024"
author: "John Kathena, Jim Ianelli"
date: today
prefer-html: true
vignette: >

  %\VignetteIndexEntry{Model update}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


## Example model runs

Below are examples which show how models can be run based on the directory location.

### Model descriptions

The following table was developed based on testing the model with different assumptions and data sources.
Key differences from the 2023 assessment configuration was the assumption that 
model estimation of variance terms was appropriate. This feature resulted in unacceptable 
residual patterns and essentially a complete down weighting of the index data.
We used the assumed variance terms (CVs) for the indices in all of the following 
model configurations:

| Model | Description    |
|----------------------|--------------------------------------------------|
| Basecase (bc)  | Model with survey "minus group" to be ages 0, and 1 instead of 0, 1, and 2 as done in the past, steepness fixed at 0.7, q estimated,  |
| m1  | As base case but with survey catchability fixed at 1.0 |
| m2  | As base case but with survey catchability fixed at 0.5 |
| m3  | As base case but with natural mortality estimated   |
| m4  | As 1.03 but with more ages in the SRR and steepness estimated with prior, sigmaR=0.6 |
| m4  | As 1.03 but with more ages in the SRR and steepness estimated with prior, sigmaR=0.6 |
| m5  | As 1.03 but with more ages in the SRR and steepness estimated with prior, sigmaR=0.6 |
| m6  | As 1.03 but with more ages in the SRR and steepness estimated with prior, sigmaR=0.6 |

```{r startup}
#| echo: false
#| output: false

library(NamibianHake)
## basic example code
library(here)
library(tidyverse)
library(ggridges)
theme_set(ggthemes::theme_few())

```

```{r getmods}
#| echo: false
#| output: false
bc <- run_nh("bc",runit=FALSE)
m1 <- run_nh("m1",runit=FALSE)
m2 <- run_nh("m2",runit=FALSE)
m3 <- run_nh("m3",runit=FALSE)
mods <-  rbind(
    read_csv(here("mods", "bc", "nh_out.csv")) |> mutate(Model = "Base Case"),
    read_csv(here("mods", "m1", "nh_out.csv")) |> mutate(Model = "Model 1"),
    read_csv(here("mods", "m2", "nh_out.csv")) |> mutate(Model = "Model 2"),
    read_csv(here("mods", "m3", "nh_out.csv")) |> mutate(Model = "Model 3")
    )
  
df <- rbind(
    data.frame(SSB = bc$SSB, R = bc$Pred_Rec, Model = "Base Case"),
    data.frame(SSB = m1$SSB, R = m1$Pred_Rec, Model = "Model 1"),
    data.frame(SSB = m2$SSB, R = m2$Pred_Rec, Model = "Model 2"),
    data.frame(SSB = m3$SSB, R = m3$Pred_Rec, Model = "Model 3") 
  ) 
```

```{r plots}
#| echo: false
#| fig.cap: "Model runs"
p1 <- df %>% ggplot(aes(x = SSB, y = R, color = Model, shape=Model)) +
  geom_point() +
  geom_line();p1
PlotAgeFit(x=bc, title="Base case",type="fishery",fage=2,lage=7)
PlotAgeFit(x=bc, title="Base case",type="survey1",fage=1,lage=7)
M<- data.frame(Year = 1964:2023,Sel = bc$S[1:60,], Model = "h=0.7, base case")
plot_sel()

```
