---
title: "Namibian hake model update, 2026"
author: "John Kathena, Jim Ianelli"
date: today
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

assessment_year <- 2026L
previous_year <- assessment_year - 1L
gradient_tolerance <- 0.001

source(here::here("R", "read-admb2.R"))
source(here::here("R", "assessment-validation.R"))
source(here::here("R", "get_relbiom.R"))

library(dplyr)
library(ggridges)
library(ggplot2)
library(knitr)
library(readr)
library(tidyr)

theme_set(ggthemes::theme_few())

noaa_colors <- c(
  navy = "#003087",
  light_blue = "#CFE8F3"
)

selectivity_ridge_data <- function(model) {
  years <- model$Year
  selectivity <- model$S[seq_along(years), , drop = FALSE]
  colnames(selectivity) <- seq.int(0L, ncol(selectivity) - 1L)

  as.data.frame(selectivity) |>
    mutate(Year = years) |>
    pivot_longer(-Year, names_to = "Age", values_to = "selectivity") |>
    group_by(Year) |>
    mutate(
      relative_selectivity = selectivity / max(selectivity, na.rm = TRUE)
    ) |>
    ungroup() |>
    mutate(
      Age = as.integer(Age),
      Year = factor(Year, levels = sort(unique(Year), decreasing = TRUE))
    )
}

plot_selectivity_ridges <- function(model) {
  selectivity_ridge_data(model) |>
    ggplot(aes(
      x = Age,
      y = Year,
      height = relative_selectivity,
      fill = relative_selectivity
    )) +
    geom_ridgeline_gradient(
      scale = 3,
      rel_min_height = 0.01,
      linewidth = 0.2
    ) +
    scale_fill_gradient(
      low = unname(noaa_colors["light_blue"]),
      high = unname(noaa_colors["navy"]),
      limits = c(0, 1),
      name = "Relative\nselectivity"
    ) +
    scale_x_continuous(breaks = 0:8) +
    labs(x = "Age (years)", y = "Year") +
    theme(legend.position = "right")
}

model_spec <- tibble::tribble(
  ~directory, ~model, ~configuration,
  "m25", "2025 baseline", "Accepted base configuration through 2025",
  "m26", "2026 update", "Unchanged base configuration with data through 2026",
  "m26_survey_q_1", "q = 1.0", "Survey catchability fixed at 1.0",
  "m26_survey_q_0_5", "q = 0.5", "Survey catchability fixed at 0.5",
  "m26_estimated_natural_mortality", "Estimated M", "Natural mortality estimated",
  "m26_dome_selectivity", "Dome selectivity", "Dome-shaped fishery selectivity",
  "m26_steepness_0_5", "h = 0.5", "Stock-recruit steepness fixed at 0.5",
  "m26_steepness_0_9", "h = 0.9", "Stock-recruit steepness fixed at 0.9"
)

read_model_csv <- function(directory) {
  read_csv(
    here::here("mods", directory, "nh_out.csv"),
    show_col_types = FALSE
  )
}

model_output <- setNames(
  lapply(model_spec$directory, read_model_csv),
  model_spec$directory
)

diagnostics <- lapply(model_spec$directory, function(directory) {
  path <- here::here("mods", directory)
  tryCatch(
    {
      x <- read_admb_convergence(path)
      tibble(
        directory = directory,
        objective = x$objective,
        max_gradient = x$max_gradient,
        minimum_hessian_eigenvalue = x$minimum_hessian_eigenvalue,
        positive_definite_hessian = x$positive_definite_hessian,
        hessian_source = x$hessian_source
      )
    },
    error = function(e) {
      header <- readLines(file.path(path, "nh.par"), n = 1L)
      tibble(
        directory = directory,
        objective = as.numeric(sub(
          ".*Objective function value = ([^ ]+).*", "\\1", header
        )),
        max_gradient = as.numeric(sub(
          ".*Maximum gradient component = ([^ ]+).*", "\\1", header
        )),
        minimum_hessian_eigenvalue = NA_real_,
        positive_definite_hessian = FALSE,
        hessian_source = "Unavailable or failed"
      )
    }
  )
}) |>
  bind_rows() |>
  left_join(model_spec, by = "directory") |>
  mutate(
    converged = is.finite(objective) &
      max_gradient <= gradient_tolerance &
      positive_definite_hessian,
    status = if_else(converged, "Converged", "Failed diagnostic criteria")
  )

read_likelihood_component <- function(directory, label) {
  lines <- readLines(
    here::here("mods", directory, "nh.rep"),
    warn = FALSE
  )
  match <- grep(paste0("^", label, "\\s+"), lines, value = TRUE)
  if (!length(match)) {
    return(NA_real_)
  }
  as.numeric(sub(paste0("^", label, "\\s+"), "", match[[1L]]))
}

likelihood_components <- lapply(model_spec$directory, function(directory) {
  tibble(
    directory = directory,
    CPUE = read_likelihood_component(directory, "CPUE"),
    `Survey biomass` = read_likelihood_component(directory, "Survey"),
    `Fishery ages` = read_likelihood_component(directory, "CAA"),
    `Survey ages` = read_likelihood_component(directory, "CAAsurv"),
    Recruitment = read_likelihood_component(directory, "SRresidual"),
    `Selectivity penalty` = read_likelihood_component(directory, "selpen"),
    Priors = read_likelihood_component(directory, "prior"),
    Total = read_likelihood_component(directory, "overall")
  )
}) |>
  bind_rows() |>
  left_join(model_spec |> select(directory, model), by = "directory")

terminal_value <- function(directory, variable, year) {
  model_output[[directory]] |>
    filter(Variable == variable, Year == year) |>
    pull(value)
}
```

# Executive summary

## Summary of changes in assessment inputs

The accepted 2025 model configuration was extended through
`r assessment_year` without changing the base-model assumptions. The update
adds the `r assessment_year` catch, CPUE, survey biomass, and composition rows
provided in `namhakdata26.dat`. The species-specific survey series in
`CapeHakes.csv` was extended through `r assessment_year` for *Merluccius
paradoxus* and *M. capensis*.

Automated checks confirmed continuous annual ADMB matrices from 1964 through
`r assessment_year`, the expected dimensions of all weight, catch,
composition, CPUE, and survey blocks, agreement among all terminal-year
controls, and valid `54321` and `12345` end-of-file sentinels.

## Summary of results

The `r assessment_year` base model converged with an objective function of
`r round(diagnostics$objective[diagnostics$directory == "m26"], 3)`, a maximum
gradient of
`r format(diagnostics$max_gradient[diagnostics$directory == "m26"], scientific = TRUE, digits = 3)`,
and a positive-definite Hessian. Estimated terminal spawning biomass was
`r round(terminal_value("m26", "SSB", assessment_year))` and terminal spawning
biomass relative to BMSY was
`r round(terminal_value("m26", "B_Bmsy", assessment_year), 3)`.

All six sensitivity configurations met the same convergence criteria as the
base model. After the selectivity slopes were estimated on the log scale with
a first-difference smoothing penalty, the dome-shaped selectivity case
converged with a maximum gradient below 0.001 and a positive-definite Hessian.

No TAC, ABC, or OFL recommendation is made in this document. Those quantities
require the agreed Namibian management procedure and management inputs.

```{r}
#| label: tbl-summary
#| tbl-cap: "Comparison of 2025 values from the previous assessment with 2026 values from the updated assessment."

bridge_variables <- c("SSB", "Depletion", "B_Bmsy", "Catch", "RY", "Catch_RY")

bridge_table <- bind_rows(
  model_output$m25 |>
    filter(Year == previous_year, Variable %in% bridge_variables) |>
    mutate(Model = "Previous assessment (2025 value)"),
  model_output$m26 |>
    filter(Year == assessment_year, Variable %in% bridge_variables) |>
    mutate(Model = "Updated assessment (2026 value)")
) |>
  select(Model, Variable, value) |>
  pivot_wider(names_from = Model, values_from = value) |>
  mutate(
    `Change (2026 minus 2025)` =
      `Updated assessment (2026 value)` -
        `Previous assessment (2025 value)`
  ) |>
  rename(Quantity = Variable)

kable(bridge_table, digits = 3)
```

# Introduction

This operational update applies the 2025 base-model structure to information
available through `r assessment_year`. Stable model methods are not repeated
in full; the emphasis is on new inputs, bridge results, model diagnostics, and
sensitivity to the principal structural assumptions considered in the
previous assessment.

# Fishery and management history

The assessment input contains annual catches from 1964 through
`r assessment_year`. Management-history text and the final
`r assessment_year` management decision should be inserted following review
by MFMR.

`TODO: add the final 2026 TAC decision and any in-season management changes.`

# Data

```{r}
#| label: tbl-input-validation
#| tbl-cap: "Automated validation of the 2026 model inputs."

input_check <- validate_nh_inputs(here::here("mods", "m26", "nh.dat"))
input_summary <- tibble(
  Check = c(
    "Annual model years",
    "Weight-at-age dimensions",
    "Fishery composition dimensions",
    "CPUE dimensions",
    "Survey dimensions",
    "Control EOF sentinel",
    "Data EOF sentinel"
  ),
  Result = c(
    paste(range(input_check$data$years), collapse = "–"),
    paste(dim(input_check$data$weight_start), collapse = " × "),
    paste(dim(input_check$data$fishery_caa), collapse = " × "),
    paste(dim(input_check$data$cpue), collapse = " × "),
    paste(dim(input_check$data$survey), collapse = " × "),
    input_check$control$eof,
    input_check$data$eof
  )
)
kable(input_summary)
```

## Survey data

```{r}
#| label: fig-species-survey
#| fig-cap: "Species-specific survey biomass observations through 2026."

cape_hakes <- read_csv(
  here::here("mods", "data", "CapeHakes.csv"),
  show_col_types = FALSE
)

ggplot(cape_hakes, aes(year, biomass, colour = strata)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = biomass / exp(sqrt(log1p(cv^2))),
      ymax = biomass * exp(sqrt(log1p(cv^2)))
    ),
    width = 0.15,
    alpha = 0.35
  ) +
  labs(x = "Year", y = "Survey biomass", colour = "Species")
```

## Other data

No additional external data series were introduced in this update. Missing
survey observations remain represented as missing or zero according to the
existing model input convention; no values were imputed to create artificial
annual continuity.

# Analytic approach

The age-structured ADMB model and 2025 base assumptions were retained. The
control file specifies the terminal year in four locations: the model terminal
year, final selectivity period, final recruitment-estimation year, and final
Ksp period. All four are set to `r assessment_year`.

The bridge comparison changes only the data and terminal year. Sensitivities
fix survey catchability at 1.0 and 0.5, estimate natural mortality, allow
dome-shaped fishery selectivity, and fix stock-recruit steepness at 0.5 and
0.9. Convergence requires a finite objective, maximum gradient no greater
than `r gradient_tolerance`, and a positive-definite ADMB Hessian.

# Results

## Bridge comparison

```{r}
#| label: fig-bridge-ssb
#| fig-cap: "Spawning biomass from the unchanged 2025 model and the 2026 data update."

bridge_series <- bind_rows(
  model_output$m25 |> mutate(Model = "2025 baseline"),
  model_output$m26 |> mutate(Model = "2026 update")
) |>
  filter(Variable == "SSB")

ggplot(bridge_series, aes(Year, value, colour = Model, fill = Model)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.15, colour = NA) +
  geom_line() +
  labs(x = "Year", y = "Spawning biomass")
```

```{r}
#| label: fig-bridge-depletion
#| fig-cap: "Estimated depletion from the 2025 baseline and 2026 update."

bridge_series |>
  select(-Variable) |>
  bind_rows(
    model_output$m25 |>
      filter(Variable == "Depletion") |>
      mutate(Model = "2025 baseline"),
    model_output$m26 |>
      filter(Variable == "Depletion") |>
      mutate(Model = "2026 update")
  ) |>
  filter(Year >= 1990, !is.na(Variable)) |>
  ggplot(aes(Year, value, colour = Model, fill = Model)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.15, colour = NA) +
  geom_line() +
  labs(x = "Year", y = "Spawning biomass / unfished spawning biomass")
```

## Time-varying fishery selectivity

The following figures use the year-normalized gradient ridge-plot presentation
from the pollock model to compare time-varying fishery selectivity from the
converged 2026 base model with the dome-shaped sensitivity. The same colour
scale is used in both figures.

For the dome-shaped sensitivity, the descending-limb slope in selectivity
period \(p\) was reparameterized as

$$
d_p = \exp(\eta_p),
$$

where $\eta_p$ is the estimated `log_SelSlope` and $d_p$ is the positive
descending slope used by the model. For ages at or above the specified onset
age $a_d$, the ascending logistic selectivity $L_p(a)$ is modified as

$$
\widetilde S_p(a) =
L_p(a)\exp\{-d_p(a-a_d)\}, \qquad a \geq a_d,
$$

and the complete age-specific curve is subsequently normalized to have a
maximum of one. Estimating $\eta_p$, rather than directly estimating a
bounded $d_p$, guarantees a positive descending slope and removes numerical
behavior associated with a slope estimate approaching a boundary.

Two penalties stabilize the period-specific estimates. The existing
shrinkage penalty controls their overall magnitude,

$$
P_{\mathrm{level}} =
12.5\sum_{p=1}^{P}(d_p-0.5)^2,
$$

while a first-difference penalty on the log slopes discourages abrupt changes
between adjacent selectivity periods,

$$
P_{\mathrm{smooth}} =
\frac{1}{2}\sum_{p=2}^{P}
\left(\frac{\eta_p-\eta_{p-1}}{0.15}\right)^2.
$$

These modifications retain time variation and permit dome-shaped curves, but
reduce confounding among poorly informed period-specific slopes and the other
parameters controlling the location and width of selectivity. In the final
run, the regularized dome-shaped model had a maximum gradient of
`r format(diagnostics$max_gradient[diagnostics$directory == "m26_dome_selectivity"], scientific = TRUE, digits = 3)`
and a positive-definite Hessian.

\clearpage

```{r}
#| label: fig-selectivity-base
#| fig-cap: "Time-varying fishery selectivity for the converged 2026 base model."
#| fig-width: 6
#| fig-height: 7.5
#| fig-pos: H

base_selectivity <- read_rep(
  here::here("mods", "m26", "nh_R.rep")
)
plot_selectivity_ridges(base_selectivity)
```

\clearpage

```{r}
#| label: fig-selectivity-dome
#| fig-cap: "Time-varying fishery selectivity for the converged 2026 dome-shaped sensitivity."
#| fig-width: 6
#| fig-height: 7.5
#| fig-pos: H

dome_selectivity <- read_rep(
  here::here("mods", "m26_dome_selectivity", "nh_R.rep")
)
plot_selectivity_ridges(dome_selectivity)
```

{{< pagebreak >}}

## Model diagnostics and sensitivities

```{r}
#| label: tbl-convergence
#| tbl-cap: "Convergence diagnostics for the bridge and sensitivity models."
#| tbl-pos: H

diagnostics |>
  select(
    Model = model,
    Objective = objective,
    `Maximum gradient` = max_gradient,
    Status = status
  ) |>
  kable(digits = 5)
```

```{r}
#| label: tbl-likelihood-components
#| tbl-cap: "Penalized negative-log-likelihood components by model. Lower values indicate a smaller contribution to the objective, but components should be compared within rather than across data types."
#| tbl-pos: H

likelihood_components |>
  mutate(model = recode(
    model,
    `2025 baseline` = "2025",
    `2026 update` = "2026",
    `Estimated M` = "Est. M",
    `Dome selectivity` = "Dome"
  )) |>
  select(
    model,
    CPUE,
    `Survey B` = `Survey biomass`,
    `Fishery CAA` = `Fishery ages`,
    `Survey CAA` = `Survey ages`,
    Recruit = Recruitment,
    `Sel. penalty` = `Selectivity penalty`,
    Priors,
    Total
  ) |>
  pivot_longer(-model, names_to = "Component", values_to = "Value") |>
  pivot_wider(names_from = model, values_from = Value) |>
  kable(digits = 2)
```

The one-year-old biomass index is not shown as an objective component because
its likelihood weight is zero in these runs. The negative component values
arise from the likelihood formulations and their estimated variance terms;
they do not indicate negative lack of fit. Totals are penalized objectives,
so their absolute values are most directly comparable for configurations
fitted to the same observations with the same penalty structure.

The notably lower total for the \(h=0.5\) model is almost entirely attributable
to the recruitment-residual component. Its recruitment contribution is
`r round(likelihood_components$Recruitment[likelihood_components$directory == "m26_steepness_0_5"], 2)`,
compared with
`r round(likelihood_components$Recruitment[likelihood_components$directory == "m26"], 2)`
for the 2026 base model: a reduction of
`r round(likelihood_components$Recruitment[likelihood_components$directory == "m26"] - likelihood_components$Recruitment[likelihood_components$directory == "m26_steepness_0_5"], 2)`
objective units. The full objective decreases by
`r round(likelihood_components$Total[likelihood_components$directory == "m26"] - likelihood_components$Total[likelihood_components$directory == "m26_steepness_0_5"], 2)`
units, while the CPUE, survey, age-composition, selectivity-penalty, and prior
components change comparatively little. With steepness fixed at 0.5, the
stock--recruit curve requires smaller estimated recruitment deviations under
the model's quadratic residual penalty. Thus the lower total primarily
reflects better agreement with the assumed stock--recruit relationship, not a
broad improvement in fit to every observed data type or, by itself, evidence
that \(h=0.5\) should replace the base assumption.

```{r}
#| label: fig-sensitivity
#| fig-cap: "Terminal spawning biomass for the converged 2026 models."
#| fig-height: 4
#| fig-pos: H

converged_directories <- diagnostics |>
  filter(converged, directory != "m25") |>
  pull(directory)

sensitivity_terminal <- lapply(converged_directories, function(directory) {
  model_output[[directory]] |>
    filter(Year == assessment_year, Variable == "SSB") |>
    mutate(directory = directory)
}) |>
  bind_rows() |>
  left_join(model_spec, by = "directory")

ggplot(
  sensitivity_terminal,
  aes(reorder(model, value), value, ymin = ymin, ymax = ymax)
) +
  geom_errorbar(width = 0.2) +
  geom_point(size = 2) +
  coord_flip() +
  labs(x = NULL, y = "Terminal spawning biomass")
```

# Control rule application

## Modeling Namibian hake survey data by species

Fisheries stock assessments require data that are reliably collected and
compiled. Assessment models should also be configured to match the assumptions
associated with the observed data. To account for survey trends between the
two hake species, the estimated observation errors were applied in a
state-space random-walk model. The observation model applies the
observation-error variances $\sigma_{j,t}^2$ for species $j$ in year $t$.
The observed indices $Z_{j,t}$ are fit to latent population trends
$\hat{Z}_{j,t}$:

$$
\log Z_{j,t} = \log \hat{Z}_{j,t} + \epsilon_{j,t},
\qquad \epsilon_{j,t} \sim N(0,\sigma_{j,t}^2).
$$

The state equation and associated process-error variance $\tau_j^2$ is:

$$
\log \hat{Z}_{j,t+1} = \log \hat{Z}_{j,t} + \eta_{j,t},
\qquad \eta_{j,t} \sim N(0,\tau_j^2).
$$

The process-error variances may be shared or allowed to vary among indices,
and the unobserved species population is estimated as a series of random
effects. The model is fit by maximum likelihood in TMB using the
[REMA package](https://afsc-assessments.github.io/rema/). Species-specific
survey observations through `r assessment_year` were fit using their supplied
CVs as observation-error inputs.

## Application to the Namibian hake stocks

The control rule was first applied using survey data and the relative
proportions of the two hake species. A more robust formulation responds
directly to the *M. paradoxus* biomass signal, independently of *M. capensis*.
For this purpose, the survey smoother was fit to *M. paradoxus* alone. The
long-term reference level is the geometric mean of predicted biomass from
1990 through `r assessment_year`.

@fig-paradoxus-biomass updates Figure 18 from the 2025 assessment with
the 2026 *M. paradoxus* survey observation.

```{r}
#| label: fig-paradoxus-biomass
#| fig-cap: "Historical REMA-smoothed biomass estimates for M. paradoxus and the geometric mean over 1990–2026."
#| fig-width: 6
#| fig-height: 5

invisible(capture.output(
  paradoxus_biomass <- get_relative_biomass(
    spp = "paradoxus",
    endyr = assessment_year
  )
))

paradoxus_series <- paradoxus_biomass$biomass_by_strata |>
  filter(strata == "paradoxus")

mean_paradoxus <- exp(mean(paradoxus_series$log_pred))
current_paradoxus_ratio <- paradoxus_series |>
  filter(year == assessment_year) |>
  summarise(value = pred / mean_paradoxus) |>
  pull(value)

ggplot(
  paradoxus_series,
  aes(
    x = year,
    y = obs,
    ymin = pred_lci,
    ymax = pred_uci
  )
) +
  geom_ribbon(fill = "#2C7FB8", alpha = 0.3) +
  geom_errorbar(
    aes(ymin = obs_lci, ymax = obs_uci),
    width = 0.2,
    alpha = 0.4
  ) +
  geom_line(aes(y = pred), colour = "#045A8D") +
  geom_point(colour = "#045A8D", alpha = 0.8) +
  geom_hline(
    yintercept = mean_paradoxus,
    linetype = "dashed",
    colour = "#D95F0E"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.05)),
    limits = c(0, NA)
  ) +
  labs(x = "Year", y = "Survey biomass (kt)")
```

The historical adjustments for three values of the species reactivity
parameter $\gamma$ are shown in @fig-paradoxus-adjustment. Values above the
long-term mean are capped at one, so this component only reduces the combined
stock advice.

```{r}
#| label: fig-paradoxus-adjustment
#| fig-cap: "Historical TAC adjustment factors for alternative reactivities to M. paradoxus biomass."

paradoxus_adjustments <- paradoxus_series |>
  transmute(
    Year = year,
    Relative_biomass = pmin(pred / mean_paradoxus, 1),
    `Gamma = 0.1` = Relative_biomass^0.1,
    `Gamma = 0.25` = Relative_biomass^0.25,
    `Gamma = 0.5` = Relative_biomass^0.5
  ) |>
  pivot_longer(
    starts_with("Gamma"),
    names_to = "Reactivity",
    values_to = "Adjustment"
  )

ggplot(
  paradoxus_adjustments,
  aes(Year, Adjustment, colour = Reactivity)
) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "TAC adjustment based on M. paradoxus biomass"
  )
```

## Control rule developments

The two-species control rule combines the overall biomass trend from the
assessment with the species-specific survey signal. Its purpose is to reduce
the exploitation rate when either combined spawning biomass or
*M. paradoxus* biomass is low. The specification evaluated in the 2025 report
is retained:

$$
TAC_y =
\begin{cases}
0, & p_y < 0.3\bar{p}, \\
\delta
\left(\frac{1}{5}\sum_{i=0}^{4}RY_{y-i}\right)
f(B_y,\lambda)
\min\left(1,\frac{\tilde{p}_y}{\bar{p}}\right)^\gamma,
& p_y \geq 0.3\bar{p},
\end{cases}
$$

where:

$$
f(B_y,\lambda) =
\begin{cases}
\left(\dfrac{B_y/B_{MSY}}{0.5}\right)^\lambda,
& B_y/B_{MSY} < 0.5, \\
1, & B_y/B_{MSY} \geq 0.5.
\end{cases}
$$

The rebuilding factor $\delta$ is 0.8 when spawning biomass is below
$B_{MSY}$ and 1.0 otherwise. The $\lambda$ term responds when spawning
biomass falls below $0.5B_{MSY}$, while $\gamma$ controls the response to
the *M. paradoxus* biomass ratio.

@tbl-control-rule applies the same four $\lambda$-$\gamma$ options
shown in the 2025 report to the unchanged 2025 model and the 2026 update.

```{r}
#| label: tbl-control-rule
#| tbl-cap: "Namibian hake control-rule calculations under the options evaluated in the 2025 report."

summary_value <- function(directory, variable) {
  model_output[[directory]] |>
    filter(Variable == variable, is.na(Year)) |>
    pull(value)
}

control_options <- tibble::tribble(
  ~Option, ~lambda, ~gamma,
  "Option 1", 1.5, 0.25,
  "Option 2", 2.0, 0.25,
  "Option 3", 2.0, 0.10,
  "Option 4", 2.0, 1.00
)

control_models <- tibble::tribble(
  ~directory, ~Model,
  "m25", "2025 baseline",
  "m26", "2026 update"
)

control_rule_results <- lapply(
  seq_len(nrow(control_models)),
  function(index) {
    directory <- control_models$directory[[index]]
    b_bmsy <- summary_value(directory, "Cur_Bmsy")
    recent_ry <- summary_value(directory, "aveRY_last5")
    delta <- if (b_bmsy < 1) 0.8 else 1

    control_options |>
      mutate(
        Model = control_models$Model[[index]],
        `Recent 5-year average replacement yield` = recent_ry,
        `Current spawning biomass / BMSY` = b_bmsy,
        `2026 M. paradoxus biomass / long-term mean` =
          current_paradoxus_ratio,
        Biomass_adjustment = if (b_bmsy < 0.5) {
          (b_bmsy / 0.5)^lambda
        } else {
          1
        },
        Species_adjustment =
          pmin(current_paradoxus_ratio, 1)^gamma,
        `Calculated TAC` =
          delta * recent_ry * Biomass_adjustment * Species_adjustment
      )
  }
) |>
  bind_rows()

control_rule_table <- bind_rows(
  control_rule_results |>
    distinct(
      Model,
      `Recent 5-year average replacement yield`,
      `Current spawning biomass / BMSY`,
      `2026 M. paradoxus biomass / long-term mean`
    ) |>
    pivot_longer(
      -Model,
      names_to = "Statistic",
      values_to = "Value"
    ),
  control_rule_results |>
    transmute(
      Model,
      Statistic = paste0(
        Option,
        ": lambda = ",
        lambda,
        ", gamma = ",
        gamma
      ),
      Value = `Calculated TAC`
    )
) |>
  pivot_wider(names_from = Model, values_from = Value)

kable(control_rule_table, digits = 3)
```

The response curves retained from the 2025 report are shown in
@fig-gamma-response and @fig-lambda-response. For the 2026 update,
*M. paradoxus* predicted biomass is
`r round(current_paradoxus_ratio, 3)` of its 1990–2026 geometric mean, and the
base-model spawning biomass is
`r round(summary_value("m26", "Cur_Bmsy"), 3)` of $B_{MSY}$.

```{r}
#| label: fig-gamma-response
#| fig-cap: "TAC adjustment for alternative values of gamma and relative M. paradoxus biomass."

gamma_response <- tidyr::crossing(
  Relative_biomass = seq(0.05, 1, by = 0.05),
  Gamma = factor(c(0.1, 0.25, 0.5))
) |>
  mutate(Adjustment = Relative_biomass^as.numeric(as.character(Gamma)))

ggplot(
  gamma_response,
  aes(Relative_biomass, Adjustment, colour = Gamma)
) +
  geom_line() +
  labs(
    x = "Relative M. paradoxus biomass",
    y = "TAC adjustment"
  )
```

```{r}
#| label: fig-lambda-response
#| fig-cap: "TAC adjustment for alternative values of lambda and spawning biomass relative to BMSY."

lambda_response <- tidyr::crossing(
  Relative_biomass = seq(0.01, 1, by = 0.01),
  Lambda = factor(c(0.5, 1, 1.5, 2))
) |>
  mutate(
    Adjustment = if_else(
      Relative_biomass < 0.5,
      (Relative_biomass / 0.5)^as.numeric(as.character(Lambda)),
      1
    )
  )

ggplot(
  lambda_response,
  aes(Relative_biomass, Adjustment, colour = Lambda)
) +
  geom_line() +
  labs(
    x = "Spawning biomass relative to BMSY",
    y = "TAC adjustment"
  )
```

# Projections

A set of simple 40-year projections was produced from the 2026 base-case
model. The scenarios hold catch at 50%, 100%, or 110% of the 2026 catch;
increase or decrease catch by 10% for 10 years before returning to the 2026
catch; or set catch to the model-estimated replacement yield. Their effects on
catch and depletion are compared in @fig-projections.

```{r}
#| label: fig-projections
#| fig-cap: "Forty-year projections under catch scenarios based on the 2026 catch. C_26 is the 2026 catch; the 10-year scenarios alter catch by 10% before returning to C_26; replacement yield sets catch to the model-estimated replacement yield."
#| fig-width: 8
#| fig-height: 7
#| fig-pos: H

projection_labels <- c(
  "0.5*C_26",
  "0.75*C_26",
  "0.9*C_26",
  "C_26",
  "1.1*C_26",
  "1.5*C_26",
  "2.0*C_26",
  "10yr up 10%",
  "10yr down 10%",
  "Replacement yield"
)

projection_data <- read_csv(
  here::here("mods", "m26", "main.out"),
  col_names = c("Scenario", "Year", "Variable", "Value"),
  show_col_types = FALSE,
  trim_ws = TRUE
) |>
  filter(
    Variable %in% c("Catch", "Depletion"),
    Scenario %in% c(1, 4, 5, 8, 9, 10),
    Year >= assessment_year
  ) |>
  mutate(
    Projection = factor(
      projection_labels[Scenario],
      levels = projection_labels[c(1, 5, 9, 8, 4, 10)]
    ),
    Variable = factor(Variable, levels = c("Catch", "Depletion"))
  )

ggplot(
  projection_data,
  aes(Year, Value, colour = Projection)
) +
  geom_line() +
  geom_point(size = 1.2) +
  facet_grid(Variable ~ ., scales = "free_y") +
  labs(x = "Year", y = "Value", colour = "Projection") +
  theme(legend.position = "right")
```

# Harvest recommendations

The ADMB model produced replacement-yield and long-term reference-point
outputs for each converged configuration. A management recommendation is not
presented because the final choice among control-rule options and other
management inputs was not supplied as part of this update. The calculations
in @tbl-control-rule reproduce the options evaluated previously and should not
be interpreted as adopted TAC advice.

# Risk table and recommendation

The 2026 base model passed the declared numerical diagnostics. Structural
uncertainty remains material: estimated natural mortality and fixed survey
catchability change the inferred scale of the stock. The dome-shaped
selectivity case now passes the numerical criteria, but it remains a structural
sensitivity rather than the selected base configuration.

# Status determination

The base-model estimate of spawning biomass relative to BMSY is reported in
@tbl-summary. Formal legal or management status terminology should be supplied
by MFMR using the applicable Namibian framework.

# Ecosystem considerations

No new ecosystem covariates were incorporated in the assessment likelihood.
The species-specific survey series is retained for evaluating the relative
signals for *M. paradoxus* and *M. capensis*.

# Data gaps and research priorities

1. Evaluate the biological plausibility and stability of the regularized
   dome-shaped selectivity alternative before using it for management
   inference.
2. Document the provenance and preprocessing of each annual input block.
3. Add retrospective analyses and likelihood profiles to the operational
   workflow.
5. Archive the final management-procedure inputs alongside the accepted
   assessment.

# Acknowledgements

The assessment update was prepared by John Kathena and Jim Ianelli with data
provided by the Namibian Ministry of Fisheries and Marine Resources.

# Auxiliary files

The assessment source, annual controls, automated validation tests, ADMB
template, and model outputs are maintained in the `NamibianHake` repository.
